---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  message = FALSE
)
```



# {langevinSSM} 

####  Habitat-driven Langevin Diffusion with Spatial Uncertainty

`{langevinSSM}` is an R package for simulating and fitting the habitat-driven Langevin diffusion to animal tracking data subject to location measurement error and temporal irregularity. The habitat-driven Langevin diffusion can provide inferences about habitat selection and utilization distributions. The package provides tools for simulating animal movement paths (`simLangevin`) and fitting the Langevin diffusion model to observed tracking data (`fitLangevin`). Location measurement error can take the form of either (older) Argos Least Squares-based locations or (newer) Argos Kalman Filter-based locations with error ellipse information. The Langevin diffusion is a continuous-time model in state-space form that estimates the underlying movement process while accounting for location measurement error and associated uncertainty in the spatial (habitat) covariates. Template Model Builder {TMB} is used for fast estimation. 

## Installation 
One can install the `langevinSSM` package from CRAN using the following command:
```{r, eval = FALSE}
install.packages("langevinSSM") 
``` 
Alternatively, one can install the package from GitHub using the `remotes` package:
```{r, eval = FALSE}
remotes::install_github("bmcclintock/langevinSSM")
```

## Usage
To simulate animal movement paths using the Langevin diffusion model, one can use the `simLangevin` function. For example:
```{r,eval=TRUE, echo=c(-10,-19)}
library(langevinSSM)
library(ggplot2)
library(terra)
library(patchwork)

# Simulate an underdamped Langevin diffusion path

par <- list(beta = c(-4, 6, 5, -0.1), # habitat selection coefficients
            sigma = 5, # diffusion (or speed) parameter
            gamma = 0.5) # autocorrelation parameter

set.seed(1, kind="Mersenne-Twister", normal.kind="Inversion")
## exampleCovs is a list of four spatial covariates (e.g., habitat features) that loads with the package
simDat <- simLangevin(model = "underdamped",
                      par = par,
                      spatialCovs = exampleCovs,
                      nbAnimals = 3)

head(simDat)

# Simulate an underdamped Langevin diffusion path with measurement error
measurementError <- list(smaj.sd = 1.5,      # sd of semi-major axis of error ellipse
                         smin.sd = 0.75,     # sd of semi-minor axis of error ellipse
                         eor = c(0,180)) # range of ellipse orientation (in degrees from north)

set.seed(1, kind="Mersenne-Twister", normal.kind="Inversion")
exampleDat <- simLangevin(model = "underdamped",
                         par = par,
                         spatialCovs = exampleCovs,
                         nbAnimals = 3,
                         obsPerAnimal = 500,
                         measurementError = measurementError)

head(exampleDat)
```                                                 
To fit the Langevin diffusion model to observed tracking data, one can use the `formatData` and `fitLangevin` functions. For example:
```{r,eval=TRUE}
# unformatDat is example data appropriate for formatData that loads with the package
head(unformatDat)

# format the data for fitLangevin
exampleDat <- formatData(unformatDat, time.unit = "hours")

head(exampleDat)

# Fit the underdamped Langevin diffusion model to simulated data with measurement error
## setting calcOSA = TRUE will calculate one-step-ahead residuals for model diagnostics
fit <- fitLangevin(model = "underdamped",
                   data = exampleDat,
                   spatialCovs = exampleCovs,
                   silent = TRUE,
                   calcOSA = TRUE)  

fit

# calculate the estimated UD
UD <- getUD(spatialCovs = exampleCovs, beta = fit$estimates$natural[1:length(exampleCovs)])

# plot the estimated (log) UD with the observed and estimated locations
plot(fit, spatialCovs = exampleCovs, data = exampleDat)

# plot residuals to check model fit
p <- plotResiduals(fit)
p$qq_x + p$qq_y + p$acf_x + p$acf_y + plot_layout(ncol=2)

# calculate the true utiliziation distribution
trueUD <- getUD(spatialCovs = exampleCovs, beta = par$beta)

# calculate similarity of true and estimated UDs using Bhattacharyya's affinity
rasterOverlap(exp(UD), exp(trueUD))
```
## Citation

If you use `{langevinSSM}` in your research, please cite it as follows:

```{r citation, echo = FALSE, comment = ""}
print(citation("langevinSSM"), bibtex = TRUE)
