---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  message = FALSE
)
```



# {langevinSSM} 

####  Habitat-driven Langevin Diffusion with Spatial Uncertainty

`{langevinSSM}` is an R package for simulating and fitting the habitat-driven Langevin diffusion to animal tracking data subject to location measurement error and temporal irregularity. The habitat-driven Langevin diffusion provides inferences about both habitat selection and utilization distributions. The package provides tools for simulating animal movement paths (`simLangevin`) and fitting the Langevin diffusion model to observed tracking data (`fitLangevin`). Location measurement error can take the form of  (older) Argos Least Squares-based locations, (newer) Argos Kalman Filter-based locations with error ellipse information, or general x- and y-axis errors (e.g. for GPS data). The Langevin diffusion is a continuous-time model in state-space form that estimates the underlying movement process while accounting for location measurement error and associated uncertainty in the spatial (habitat) covariates. Template Model Builder {TMB} is used for fast estimation. 

## Installation 
One can install the `langevinSSM` package from CRAN using the following command:
```{r, eval = FALSE}
install.packages("langevinSSM") 
``` 
Alternatively, one can install the package from GitHub using the `remotes` package:
```{r, eval = FALSE}
remotes::install_github("bmcclintock/langevinSSM")
```

## Usage
To simulate animal movement paths using the Langevin diffusion model, one can use the `simLangevin` function. For example:
```{r sim,eval=TRUE, echo=c(-14,-22)}
library(langevinSSM)
library(ggplot2)
library(terra)
library(patchwork)

# Simulate an underdamped Langevin diffusion path

par <- list(beta = c(-4, 6, 5, -0.1), # habitat selection coefficients
            sigma = 5, # diffusion (or speed) parameter
            gamma = 0.5) # autocorrelation parameter

# calculate the true utiliziation distribution
## exampleCovs is a list of four spatial covariates (e.g., habitat features) that loads with the package
trueUD <- getUD(spatialCovs = exampleCovs, beta = par$beta)

set.seed(1, kind="Mersenne-Twister", normal.kind="Inversion")
simDat <- simLangevin(model = "underdamped",
                      par = par,
                      spatialCovs = exampleCovs,
                      nbAnimals = 3)

head(simDat)

# Simulate an underdamped Langevin diffusion path with measurement error
measurementError <- list(smaj.sd = 1.5,      # sd of semi-major axis of error ellipse
                         smin.sd = 0.75,     # sd of semi-minor axis of error ellipse
                         eor.lim = c(0,180)) # range of ellipse orientation (in degrees from north)

set.seed(1, kind="Mersenne-Twister", normal.kind="Inversion")
exampleDat <- simLangevin(model = "underdamped",
                         par = par,
                         spatialCovs = exampleCovs,
                         nbAnimals = 3,
                         obsPerAnimal = 500,
                         measurementError = measurementError)

head(exampleDat)
```                                                 

To fit the Langevin diffusion model to observed tracking data, one can use the `formatData` and `fitLangevin` functions. For example:
```{r fit,eval=TRUE}
# unformatDat is example data appropriate for formatData that loads with the package
head(unformatDat)

# format the data for fitLangevin
exampleDat <- formatData(unformatDat, time.unit = "hours")

head(exampleDat)

# Fit the overdamped Langevin diffusion model to simulated data with measurement error
fit_over <- fitLangevin(model = "overdamped",
                   data = exampleDat,
                   spatialCovs = exampleCovs,
                   silent = TRUE)  

fit_over

# Fit the underdamped Langevin diffusion model
fit_under <- fitLangevin(model = "underdamped",
                   data = exampleDat,
                   spatialCovs = exampleCovs,
                   silent = TRUE)  

fit_under
```
### Post-processing functions
#### Utilization distribution
```{r ud,eval=TRUE}
# calculate the estimated UD
UD <- getUD(spatialCovs = exampleCovs, fit = fit_under)

# plot the estimated (log) UD with the observed and estimated locations
plot(fit_under, spatialCovs = exampleCovs, data = exampleDat)
```

#### Other S3 methods for `fitLangevin` objects
```{r s3,eval=TRUE}
coef(fit_under) # fixed effect estimates

confint(fit_under) # confidence intervals for fixed effects

mu_ci <- confint(fit_under, type= "mu") # confidence intervals for true locations
head(mu_ci)

AIC(fit_under) # for comparing models with different fixed effects

BIC(fit_under) # for comparing models with different fixed effects
```
#### One-step-ahead residuals
```{r osa,eval=TRUE,cache=TRUE}
# calculate one-step-ahead residuals for model diagnostics
## can also be used to compare "underdamped" vs "overdamped" model fits

res_under <- residuals(fit_under, data = exampleDat, spatialCovs = exampleCovs, ncores = 3)
res_under

# plot residuals to check model fit
p_under <- plot(res_under)
p_under$qq_x + p_under$qq_y + p_under$acf_x + p_under$acf_y + plot_layout(ncol=2)

res_over <- residuals(fit_over, data = exampleDat, spatialCovs = exampleCovs, ncores = 3)
res_over

p_over <- plot(res_over)
p_over$qq_x + p_over$qq_y + p_over$acf_x + p_over$acf_y + plot_layout(ncol=2)
```

#### Bhattacharyya's affinity
```{r ba,eval=TRUE}
# calculate similarity of true and estimated UDs using Bhattacharyya's affinity
rasterOverlap(exp(UD), exp(trueUD))
```
#### Regional presence probability
```{r regionProb,eval=TRUE}
# calculate probability of being in specified region
## create a spatial mask for the region of interest
d2c <- exampleCovs$d2c < 2.5
reg_prob <- regionProb(fit_under,
                       spatialCovs = exampleCovs, 
                       mask = d2c, # region of interest
                       nSims = 1000, # number of Monte Carlo simulations
                       show_progress = FALSE)

reg_prob

plot(reg_prob, log = TRUE)
```

## Citation

If you use `{langevinSSM}` in your research, please cite it as follows:

```{r citation, echo = FALSE, comment = ""}
print(citation("langevinSSM"), bibtex = TRUE)
